Lineage Analysis of CRISPR and shRNA Screens

### 2018-08-21 lineages_18Q3.Rmd
add CRISPR heatmap code; save data matrices for heatmap

  # Mutation status
  crispr_color <- as.character(gene$Color)
  names(crispr_color) <- gene$Mutation_Status
  sig <- filter(crispr_signif, Hugo_Symbol == g)
  plot_mut <- ggplot(data = gene, aes(x = Mutation_Status, y = Score, color = Mutation_Status)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(alpha = 0.3, size = 0.7, position = position_jitter(w = 0.05)) +
    scale_color_manual(values = crispr_color) +
    geom_hline(yintercept = 0, linetype = 2, lwd = 0.3) +
    theme_light() +
    theme(legend.position = "none") +
    labs(x = paste0(g, " Mutation Status"), y = "CERES Score",
         title = "Mutation Status",
         subtitle = paste0("Wilcoxon test:\n\tBH-corrected p-value: ", sig$p.adj.short, "\n\tUncorrected p-value: ", sig$p.short))

@@

# Melt
cn_melt <- melt(data = cn, id.vars = "Broad_ID", measure.vars = colnames(cn[2:ncol(cn)]))
colnames(cn_melt) <- c("Broad_ID", "Hugo_Symbol", "Copy_Number")

@@

Gene expression data (Reads Per Kilobase of transcript, per Million mapped reads, RPKM).
```{r}
ge <- read.delim("./../CCLE_DepMap_18q3_RNAseq_RPKM_20180718.gct.gz", skip = 2, header = TRUE, sep = "\t", check.names = FALSE)
```

```{r, eval=FALSE}
# Edit columns
ge$Name <- NULL
colnames(ge)[1] <- "Hugo_Symbol"

# Melt
ge_melt <- melt(data = ge, id.vars = "Hugo_Symbol", measure.vars = colnames(ge[2:ncol(ge)]))
colnames(ge_melt)[3] <- "RPKM"
## Split variable column
ge_melt <- with(ge_melt, cbind(Hugo_Symbol, colsplit(variable, pattern = " ", names = c("CCLE_Name", "Broad_ID")), RPKM))
## Remove parentheses around Broad IDs
ge_melt$Broad_ID <- gsub("\\(|\\)", "", ge_melt$Broad_ID)

saveRDS(ge_melt, "./../ge_melt_18Q3.rds", compress = "xz")
```

```{r}
ge_melt <- readRDS("./../ge_melt_18Q3.rds")
```

@@

Merge annotation data:
```{r, eval=FALSE}
# Melt CRISPR dataset for merging
crispr_melt <- melt(crispr, id.vars = "Broad_ID", measure.vars = colnames(crispr)[2:ncol(crispr)])
colnames(crispr_melt) <- c("Broad_ID", "Hugo_Symbol", "Score")

@@

# Melt shRNA dataset for merging
shrna_melt <- melt(shrna , id.vars = "Hugo_Symbol", measure.vars = colnames(shrna)[2:ncol(shrna)])
colnames(shrna_melt) <- c("Hugo_Symbol", "CCLE_Name", "Score")
shrna_melt <- drop_na(shrna_melt)
```

@@

saveRDS(shrna_data, "./../shrna_data_18Q3.rds", compress = "xz")
```

```{r}
shrna_data <- readRDS("./../shrna_data_18Q3.rds")
shrna_ccl <- data.frame("Broad_ID" = shrna_data$Broad_ID)
```

@@

knitr::kable(crispr_signif[, c(1, 5, 6, 7, 8, 10)], caption = "CRISPR Screen: Wilcoxon Test Results Comparing Mutant and Wildtype Cell Lines (Benjamini-Hochberg-corrected p-values: * p <= 0.05, ** p <= 0.01, *** p <= 0.001, **** p <= 0.0001)") %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% footnote(general = "Wilcoxon test Benjamini-Hochberg-corrected p-values:\n*: p <= 0.05\n**: p <= 0.01\n***: p <= 0.001\n****: p <= 0.0001") %>% scroll_box(width = "900px", height = "450px")
```

@@

knitr::kable(shrna_signif[, c(1, 5, 6, 7, 8, 10)], caption = "shRNA Screen: Wilcoxon Test Results Comparing Mutant and Wildtype Cell Lines (Benjamini-Hochberg-corrected p-values: * p <= 0.05, ** p <= 0.01, *** p <= 0.001, **** p <= 0.0001)") %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% footnote(general = "Wilcoxon test Benjamini-Hochberg-corrected p-values:\n*: p <= 0.05\n**: p <= 0.01\n***: p <= 0.001\n****: p <= 0.0001") %>% scroll_box(width = "900px", height = "450px")
```

@@

```{r, eval=FALSE}
shrna_signif_lineage <- compare_means(Score ~ Mutation_Status, group.by = c("Hugo_Symbol", "lineage_name"), data = shrna_data, method = "wilcox.test", p.adjust.method = "BH")
shrna_signif_lineage <- adj_signif(shrna_signif_lineage)
shrna_signif_lineage <- mutate(shrna_signif_lineage, lineage_name = reorder(lineage_name, p.adj, mean))
saveRDS(shrna_signif_lineage, "./../shrna_signif_lineage.rds")
```

```{r, eval=FALSE}
shrna_signif_lineage <- readRDS("./../shrna_signif_lineage.rds")
```

@@

```{r, eval=FALSE}
# Set up heatmap data
chm_data <- t(crispr)
colnames(chm_data) <- chm_data[1, ]
chm_data <- chm_data[-1, ]
chm_cols <- colnames(chm_data)
chm_data <- apply(chm_data, 2, as.double)
chm_data <- apply(chm_data, 2, function(x) (x - mean(x)) / var(x))
colnames(chm_data) <- chm_cols

# Metadata anotation dataframe
chm_annot <- merge(merge(data.frame("Broad_ID" = crispr$Broad_ID), crispr_meta[, c("Broad_ID", "cell_line_SSMD", "cas9_activity", "culture_type", "primary_tissue")], by = "Broad_ID", all.x = TRUE), ccl_info[, c("Broad_ID", "Primary.Disease", "Gender", "Source")], by = "Broad_ID", all.x = TRUE)
rownames(chm_annot) <- chm_annot$Broad_ID
chm_annot$Broad_ID <- NULL
colnames(chm_annot) <- c("Cell_Line_SSMD", "Cas9_Activity", "Culture_Type", "Primary_Tissue", "Primary_Disease", "Gender", "Source")
chm_annot$Cell_Line_SSMD <- as.numeric(chm_annot$Cell_Line_SSMD)
chm_annot$Cas9_Activity <- as.numeric(chm_annot$Cas9_Activity)

@@

# Gene expression
ge_chm_grid <- merge(crispr_grid, ge_melt[, c("Hugo_Symbol", "Broad_ID", "RPKM")], by = c("Broad_ID", "Hugo_Symbol"), all.x = TRUE)
```

```{r, eval=FALSE}
chma <- HeatmapAnnotation(df = chm_annot)

chm_euc <- Heatmap(chm_data, name = "CERES Score",
        bottom_annotation = chma,
        bottom_annotation_height = unit(5, "in"),
        row_title = "Genes", column_title = "Cancer Cell Lines (Broad IDs)",
        col = colorRamp2(c(-15, 0, 5), c("purple4", "white", "seagreen4")),
        column_title_gp = gpar(fontsize = 60, fontface = "bold"),
        show_column_names = FALSE,
        row_title_gp = gpar(fontsize = 60, fontface = "bold"),
        show_row_names = FALSE,
        row_dend_reorder = TRUE,
        column_dend_reorder = TRUE,
        clustering_distance_rows = "euclidean",
        clustering_distance_columns = "euclidean",
        clustering_method_rows = "complete",
        clustering_method_columns = "complete",
        column_dend_height = unit(4, "in"),
        row_dend_width = unit(4, "in"),
        width = 3)

pdf("./plots_18Q3/crispr_heatmap_euclidean.pdf", width = 100, height = 100, paper = "special") 
draw(chm_euc, heatmap_legend_side = "right", annotation_legend_side = "right")
seekViewport("annotation_Cell_Line_SSMD")
grid.text("Cell Line SSMD", unit(1, "npc") + unit(2, "mm"), 0.5, gp = gpar(fontsize = 30), default.units = "npc", just = "left")
seekViewport("annotation_Cas9_Activity")
grid.text("Cas9 Activity", unit(1, "npc") + unit(2, "mm"), 0.5, gp = gpar(fontsize = 30), default.units = "npc", just = "left")
seekViewport("annotation_Culture_Type")
grid.text("Culture Type", unit(1, "npc") + unit(2, "mm"), 0.5, gp = gpar(fontsize = 30), default.units = "npc", just = "left")
seekViewport("annotation_Primary_Tissue")
grid.text("Primary Tissue", unit(1, "npc") + unit(2, "mm"), 0.5, gp = gpar(fontsize = 30), default.units = "npc", just = "left")
seekViewport("annotation_Primary_Disease")
grid.text("Primary Disease", unit(1, "npc") + unit(2, "mm"), 0.5, gp = gpar(fontsize = 30), default.units = "npc", just = "left")
seekViewport("annotation_Gender")
grid.text("Gender", unit(1, "npc") + unit(2, "mm"), 0.5, gp = gpar(fontsize = 30), default.units = "npc", just = "left")
seekViewport("annotation_Source")
grid.text("Source", unit(1, "npc") + unit(2, "mm"), 0.5, gp = gpar(fontsize = 30), default.units = "npc", just = "left")
dev.off()
```

>>>>

  # Mutation status
  crispr_color <- as.character(gene$Color)
  names(crispr_color) <- gene$Mutation_Status
  sig <- filter(crispr_signif, Hugo_Symbol == g)
  plot_mut <- ggplot(data = gene, aes(x = Mutation_Status, y = Score, color = Mutation_Status)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(alpha = 0.3, size = 0.7, position = position_jitter(w = 0.05)) +
    scale_color_manual(values = crispr_color) +
    geom_hline(yintercept = 0, linetype = 2, lwd = 0.3) +
    theme_light() +
    theme(legend.position = "none") +
    labs(x = paste0(g, " Mutation Status"), y = "CERES Score",
         title = "Mutation Status",
         subtitle = paste0("Wilcoxon test:\n- BH-corrected p-value: ", sig$p.adj.short, "\n- Uncorrected p-value: ", sig$p.short))

@@

# Melt
cn_melt <- melt(data = cn, id.vars = "Broad_ID", measure.vars = colnames(cn[2:ncol(cn)]), variable.name = "Hugo_Symbol", value.name = "Copy_Number")

@@

Gene expression data (Reads Per Kilobase of transcript, per Million mapped reads, RPKM).
```{r, eval=FALSE}
ge <- read.delim("./../crispr_lineages_giant_files/CCLE_DepMap_18q3_RNAseq_RPKM_20180718.gct.gz", skip = 2, header = TRUE, sep = "\t", check.names = FALSE)

# Edit columns
ge$Name <- NULL
colnames(ge)[1] <- "Hugo_Symbol"

# Melt
ge_melt <- melt(data = ge, id.vars = "Hugo_Symbol", measure.vars = colnames(ge[2:ncol(ge)]), value.name = "RPKM")
## Split variable column
ge_melt <- with(ge_melt, cbind(Hugo_Symbol, colsplit(variable, pattern = " ", names = c("CCLE_Name", "Broad_ID")), RPKM))
## Remove parentheses around Broad IDs
ge_melt$Broad_ID <- gsub("\\(|\\)", "", ge_melt$Broad_ID)

saveRDS(ge_melt, "./../crispr_lineages_giant_files/ge_melt_18Q3.rds", compress = "xz")
```

```{r}
ge_melt <- readRDS("./../crispr_lineages_giant_files/ge_melt_18Q3.rds")
```

@@

Merge annotation data:
```{r, eval=FALSE}
# Melt CRISPR dataset for merging
crispr_melt <- melt(crispr, id.vars = "Broad_ID", measure.vars = colnames(crispr)[2:ncol(crispr)], variable.name = "Hugo_Symbol", value.name = "Score")

@@

# Melt shRNA dataset for merging
shrna_melt <- melt(shrna , id.vars = "Hugo_Symbol", measure.vars = colnames(shrna)[2:ncol(shrna)], variable.name = "CCLE_Name", value.name = "Score")
shrna_melt <- drop_na(shrna_melt)
```

@@

saveRDS(shrna_data, "./../crispr_lineages_giant_files/shrna_data_18Q3.rds", compress = "xz")
```

```{r}
shrna_data <- readRDS("./../crispr_lineages_giant_files/shrna_data_18Q3.rds")
shrna_ccl <- data.frame("Broad_ID" = shrna_data$Broad_ID)
```

@@

knitr::kable(crispr_signif[1:100, c(1, 5, 6, 7, 8, 10)], caption = "CRISPR Screen: Wilcoxon Test Results Comparing Mutant and Wildtype Cell Lines (Benjamini-Hochberg-corrected p-values: * p <= 0.05, ** p <= 0.01, *** p <= 0.001, **** p <= 0.0001)") %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% footnote(general = "Wilcoxon test Benjamini-Hochberg-corrected p-values:\n*: p <= 0.05\n**: p <= 0.01\n***: p <= 0.001\n****: p <= 0.0001") %>% scroll_box(width = "900px", height = "450px")
```

@@

knitr::kable(shrna_signif[1:100, c(1, 5, 6, 7, 8, 10)], caption = "shRNA Screen: Wilcoxon Test Results Comparing Mutant and Wildtype Cell Lines (Benjamini-Hochberg-corrected p-values: * p <= 0.05, ** p <= 0.01, *** p <= 0.001, **** p <= 0.0001)") %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% footnote(general = "Wilcoxon test Benjamini-Hochberg-corrected p-values:\n*: p <= 0.05\n**: p <= 0.01\n***: p <= 0.001\n****: p <= 0.0001") %>% scroll_box(width = "900px", height = "450px")
```

@@

```{r, eval=FALSE}
shrna_signif_lineage <- compare_means(Score ~ Mutation_Status, group.by = c("Hugo_Symbol", "lineage_name"), data = shrna_data, method = "wilcox.test", p.adjust.method = "BH")
shrna_signif_lineage <- adj_signif(shrna_signif_lineage)
shrna_signif_lineage <- mutate(shrna_signif_lineage, lineage_name = reorder(lineage_name, p.adj, mean))
saveRDS(shrna_signif_lineage, "./../crispr_lineages_giant_files/shrna_signif_lineage.rds")
```

```{r, eval=FALSE}
shrna_signif_lineage <- readRDS("./../crispr_lineages_giant_files/shrna_signif_lineage.rds")
```

@@

Make and save heatmap data matrices:
```{r, eval=FALSE}
# Set up heatmap data
chm_scores <- t(crispr)
colnames(chm_scores) <- chm_scores[1, ]
chm_scores <- chm_scores[-1, ]
chm_cols <- colnames(chm_scores)
chm_scores <- apply(chm_scores, 2, as.double)
chm_scores <- apply(chm_scores, 2, function(x) (x - mean(x)) / var(x))
colnames(chm_scores) <- chm_cols
saveRDS(chm_scores, "./data_munging/rds/crispr_heatmap_scores.rds")

@@

# Gene expression
ge_filt <- filter(ge_melt, Hugo_Symbol %in% unique(crispr_data$Hugo_Symbol))[, c("Hugo_Symbol", "Broad_ID", "RPKM")]
chm_ge_grid <- merge(crispr_grid, ge_filt, by = c("Broad_ID", "Hugo_Symbol"), all.x = TRUE)
chm_ge <- reshape(chm_ge_grid, idvar = "Hugo_Symbol", timevar = "Broad_ID", direction = "wide")
chm_ge_rows <- chm_ge$Hugo_Symbol
chm_ge$Hugo_Symbol <- NULL
chm_ge_cols <- colnames(chm_ge) %>% gsub("RPKM.", "", .)
chm_ge <- matrix(as.numeric(unlist(chm_ge)), nrow = nrow(chm_ge))
rownames(chm_ge) <- chm_ge_rows
colnames(chm_ge) <- chm_ge_cols
saveRDS(chm_ge, "./data_munging/rds/crispr_heatmap_ge.rds")

# Mutation status
chm_mut_grid <- merge(crispr_grid, maf_df[, c("Broad_ID", "Hugo_Symbol", "Mutation_Status")], by = c("Broad_ID", "Hugo_Symbol"), all.x = TRUE)
chm_mut_grid$Mutation_Status <- ifelse(is.na(chm_mut_grid$Mutation_Status), 0, 1)
chm_mut <- reshape(chm_mut_grid, idvar = "Hugo_Symbol", timevar = "Broad_ID", direction = "wide")
chm_mut_rows <- chm_mut$Hugo_Symbol
chm_mut$Hugo_Symbol <- NULL
chm_mut_cols <- colnames(chm_mut) %>% gsub("Mutation_Status.", "", .)
chm_mut <- matrix(as.numeric(unlist(chm_mut)), nrow = nrow(chm_mut))
rownames(chm_mut) <- chm_mut_rows
colnames(chm_mut) <- chm_mut_cols
saveRDS(chm_mut, "./data_munging/rds/crispr_heatmap_mut.rds")

# Copy number
cn_filt <- filter(cn_melt, Hugo_Symbol %in% unique(crispr_data$Hugo_Symbol))
chm_cn_grid <- merge(crispr_grid, cn_melt, by = c("Broad_ID", "Hugo_Symbol"), all.x = TRUE)
chm_cn <- reshape(chm_cn_grid, idvar = "Hugo_Symbol", timevar = "Broad_ID", direction = "wide")
chm_cn_rows <- chm_cn$Hugo_Symbol
chm_cn$Hugo_Symbol <- NULL
chm_cn_cols <- colnames(chm_cn) %>% gsub("Copy_Number.", "", .)
chm_cn <- matrix(as.numeric(unlist(chm_cn)), nrow = nrow(chm_cn))
rownames(chm_cn) <- chm_cn_rows
colnames(chm_cn) <- chm_cn_cols
saveRDS(chm_cn, "./data_munging/rds/crispr_heatmap_cn.rds")
```

Load heatmap matrices:
```{r}
chm_scores <- readRDS("./data_munging/rds/crispr_heatmap_scores.rds")
chm_ge <- readRDS("./data_munging/rds/crispr_heatmap_ge.rds")
chm_mut <- readRDS("./data_munging/rds/crispr_heatmap_mut.rds")
chm_cn <- readRDS("./data_munging/rds/crispr_heatmap_cn.rds")

# Metadata anotation dataframe
chm_annot <- merge(merge(data.frame("Broad_ID" = crispr$Broad_ID), crispr_meta[, c("Broad_ID", "cell_line_SSMD", "cas9_activity", "culture_type", "primary_tissue")], by = "Broad_ID", all.x = TRUE), ccl_info[, c("Broad_ID", "Primary.Disease", "Gender", "Source")], by = "Broad_ID", all.x = TRUE)
rownames(chm_annot) <- chm_annot$Broad_ID
chm_annot$Broad_ID <- NULL
colnames(chm_annot) <- c("Cell_Line_SSMD", "Cas9_Activity", "Culture_Type", "Primary_Tissue", "Primary_Disease", "Gender", "Source")
chm_annot$Cell_Line_SSMD <- as.numeric(chm_annot$Cell_Line_SSMD)
chm_annot$Cas9_Activity <- as.numeric(chm_annot$Cas9_Activity)
```

Heatmap code:
```{r, eval=FALSE}
chm_scores_plot <- Heatmap(chm_scores, name = "CERES Score",
        bottom_annotation = HeatmapAnnotation(df = chm_annot),
        bottom_annotation_height = unit(5, "in"),
        row_title = "Genes", column_title = "Cancer Cell Lines (Broad IDs)",
        col = colorRamp2(c(-15, 0, 5), c("purple4", "white", "seagreen4")),
        na_col = "black",
        column_title_gp = gpar(fontsize = 60, fontface = "bold"),
        show_column_names = FALSE,
        row_title_gp = gpar(fontsize = 60, fontface = "bold"),
        show_row_names = FALSE,
        row_dend_reorder = TRUE,
        column_dend_reorder = TRUE,
        clustering_distance_rows = "euclidean",
        clustering_distance_columns = "euclidean",
        clustering_method_rows = "complete",
        clustering_method_columns = "complete",
        column_dend_height = unit(4, "in"),
        row_dend_width = unit(4, "in"),
        width = 3)
chm_ge_plot <- Heatmap(chm_ge, name = "Gene Expression (RPKM)",
        cluster_rows = FALSE, cluster_columns = FALSE,
        show_row_names = FALSE, show_column_names = FALSE,
        column_title = "Gene Expression",
        column_title_gp = gpar(fontsize = 60, fontface = "bold"),
        column_order = colnames(chm_scores)[unlist(column_order(chm_scores_plot))],
        row_order = rownames(chm_scores)[unlist(row_order(chm_scores_plot))],
        width = 1)
chm_cn_plot <- Heatmap(chm_cn, name = "Copy Number",
        cluster_rows = FALSE, cluster_columns = FALSE,
        show_row_names = FALSE, show_column_names = FALSE,
        column_title = "Copy Number",
        column_title_gp = gpar(fontsize = 60, fontface = "bold"),
        column_order = colnames(chm_scores)[unlist(column_order(chm_scores_plot))],
        row_order = rownames(chm_scores)[unlist(row_order(chm_scores_plot))],
        width = 1)
chm_mut_plot <- Heatmap(chm_mut, name = "Mutation Status",
        col = c("cyan3", "darkorchid"),
        cluster_rows = FALSE, cluster_columns = FALSE,
        show_row_names = FALSE, show_column_names = FALSE,
        column_title = "Mutation Status",
        column_title_gp = gpar(fontsize = 60, fontface = "bold"),
        column_order = colnames(chm_scores)[unlist(column_order(chm_scores_plot))],
        row_order = rownames(chm_scores)[unlist(row_order(chm_scores_plot))],
        width = 1)

chm_all_plot <- chm_scores_plot + chm_ge_plot + chm_cn_plot + chm_mut_plot

pdf("./plots_18Q3/crispr_heatmap_all.pdf", width = 100, height = 100, paper = "special") 
draw(chm_all_plot, heatmap_legend_side = "right", annotation_legend_side = "right")
seekViewport("annotation_Cell_Line_SSMD")
grid.text("Cell Line SSMD", unit(1, "npc") + unit(2, "mm"), 0.5, gp = gpar(fontsize = 30), default.units = "npc", just = "left")
seekViewport("annotation_Cas9_Activity")
grid.text("Cas9 Activity", unit(1, "npc") + unit(2, "mm"), 0.5, gp = gpar(fontsize = 30), default.units = "npc", just = "left")
seekViewport("annotation_Culture_Type")
grid.text("Culture Type", unit(1, "npc") + unit(2, "mm"), 0.5, gp = gpar(fontsize = 30), default.units = "npc", just = "left")
seekViewport("annotation_Primary_Tissue")
grid.text("Primary Tissue", unit(1, "npc") + unit(2, "mm"), 0.5, gp = gpar(fontsize = 30), default.units = "npc", just = "left")
seekViewport("annotation_Primary_Disease")
grid.text("Primary Disease", unit(1, "npc") + unit(2, "mm"), 0.5, gp = gpar(fontsize = 30), default.units = "npc", just = "left")
seekViewport("annotation_Gender")
grid.text("Gender", unit(1, "npc") + unit(2, "mm"), 0.5, gp = gpar(fontsize = 30), default.units = "npc", just = "left")
seekViewport("annotation_Source")
grid.text("Source", unit(1, "npc") + unit(2, "mm"), 0.5, gp = gpar(fontsize = 30), default.units = "npc", just = "left")
dev.off()
```
